Detecting Cosmic Reionization using Bi-Spectrum Phase

### 2018-04-27 RefereeReportResponse-17Apr2018.txt

We thank the referees for their valuable comments and suggestions. We
are very delighted that both reviews have been very positive. In 
addressing the comments, we find that the manuscript has gained in
quality. We have responded to specific comments (see below).

@@

----------------------------------------------------------------------
Report of Referee B -- LQ16046/Thyagarajan
----------------------------------------------------------------------

I feel this paper should be published in PRL. It proposes an elegant
solution to a challenge that has plagued interferometers attempting to
exploit the faint radio signal from neutral hydrogen gas at high
redshifts as a cosmological tool for understanding the epoch of
reionization.

Indeed, my main suggestion is to expand the scope of the paper by
mentioning in the abstract and introduction that the technique has
broader application than just to measurements of the EoR. I believe
this broader focus would increase the impact of the paper
significantly. Cosmological measurements of HI at lower redshifts (<
2) are also being pursued vigorously by a variety of radio
interferometers (CHIME, Tianlai, HIRAX) that face the same challenges
as EoR instruments such as HERA. The simulations described in the
paper using HERA as an example are equally relevant for these other
interferometers, which also must precisely calibrate instrumental
gains in the presence of strong astrophysical foregrounds. There is
considerable interest in the cosmological community to develop HI
intensity mapping as an efficient tool for mapping large-scale
structure. The hoped-for payoff is improved precision on cosmological
parameters such as the neutrino mass and the equation of state of the
dark energy. Outside of the introduction, I believe that little or
nothing would need to be changed in the rest of the paper.

[Response] Added a sentence in the Abstract, and a paragraph in the
Introduction about applicability to other 21 cm intensity mapping
experiments that will measure the BAO, for example.

### 2018-04-27 ResponseToEditor.txt

To the editorial staff:

We appreciate the referees' reports and their positive views on 
our work. We have responded to their comments, attached in a 
separate file.

We are attaching two versions of the Latex file. One is the 
updated ms.tex (ms.pdf). The other is ms_changes.tex (and 
ms_changes.pdf) and it contain all the changes highlighted in 
red.

Regarding the length of the manuscript, we believe there has 
been a mistake. In the email from PRL on 7 March 2018, the word 
count summary was stated to be:

-----------------------------------------------------------------
  Figure   Aspect Ratio   Wide?   Word Equivalent
     1         1.54         No        117
     2         1.06         No        161
     3         3.73        Yes        200

WORD COUNT SUMMARY
    Note: Text word count excludes title, abstract, byline, PACS,
          receipt date, acknowledgments, and references.

              Text word count   4003
        Equations word equiv.    192
          Figures word equiv.    478
           Tables word equiv.      0
                              ------
                       TOTAL    4673 (Maximum length is 3750)
-----------------------------------------------------------------

Subsequently, we made substantial changes to reduce the length by
at least 100 words including substantial changes to figure sizes. 
We confirmed this through "texcount" on the command line and 
carefully kept track of the word count at the time of initial 
submission (5 March 2018) as well as the trimmed one (12 March 
2018).

In the recent email from PRL on 17 April 2018, the word count 
was found by PRL to be: 
-----------------------------------------------------------------
  Figure   Aspect Ratio   Wide?   Word Equivalent
     1         1.54         No        117
     2         1.06         No        161
     3         3.73        Yes        200

WORD COUNT SUMMARY
    Note: Text word count excludes title, abstract, byline, PACS,
          receipt date, acknowledgments, and references.

              Text word count   4003
        Equations word equiv.    192
          Figures word equiv.    478
           Tables word equiv.      0
                              ------
                       TOTAL    4516 (Maximum length is 3750)
-----------------------------------------------------------------
You will notice that the word count for the text, equations, and 
figures are exactly identical between 17 April and 7 March (which
is impossible), although the total has changed from 4673 to 4516. 
Not only has the itemized count not been updated, but the new 
total is also inaccurate.

We understand the output of texcount may not be identical to the 
software PRL uses. Therefore, we request PRL to give us an 
accurate count which we can use as a reference to further trim 
the manuscript, if necessary.

In the meanwhile, we request that the editorial office pass on
the revised manuscript to the referees for a speedy acceptance.

Best regards,
The authors.

### 2018-04-27 ms_changes.tex
Update on Overleaf.

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{amssymb}   % for math
\usepackage{hyperref}% add hypertext capabilities
\usepackage{aas_macros}
\usepackage[caption=false]{subfig}
\usepackage[final]{changes}
\definechangesauthor[name={Nithyanandan Thyagarajan}, color=red]{NT}
\definechangesauthor[name={Chris Carilli}, color=blue]{CC}
\definechangesauthor[name={Bojan Nikolic}, color=green]{BN}
\setremarkmarkup{(#2)}
% \usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\begin{abstract} 
  Detecting neutral Hydrogen (H\,{\sc i}) via the 21~cm line emission from the intergalactic medium at $z\gtrsim 6$ has been identified as one of the most promising probes of the epoch of cosmic reionization -- a major phase transition of the Universe. However, these studies face severe challenges imposed by the bright foreground emission from cosmic objects. Current techniques require precise instrumental calibration to separate the weak H\,{\sc i} line signal from the foreground continuum emission. We propose to mitigate this calibration requirement by using \replaced[id=NT]{measurements of the interferometric bi-spectrum phase}{the phase of the bi-spectrum from interferometric measurements}. Bi-spectrum phase is unaffected by antenna-based direction-independent calibration errors and hence for a compact array it depends on the sky brightness distribution only (subject to the usual thermal-like noise). We show that the bi-spectrum phase of foreground synchrotron continuum has a characteristically smooth spectrum relative to the cosmological line signal. The two can be separated effectively by exploiting this spectral difference using Fourier techniques, while eliminating the need for precise antenna-based calibration of phases introduced by the instrument, and the ionosphere, inherent in existing approaches. Using fiducial models for continuum foregrounds, and for the cosmological H\,{\sc i} signal, we show the latter should be detectable in bi-spectrum phase spectra, with reasonable significance at $|k_\parallel| \gtrsim 0.5\,h$~Mpc$^{-1}$, using existing instruments. \added[id=NT]{Our approach will also benefit other H\,{\sc i} intensity mapping experiments that face similar challenges, such as those measuring Baryon Acoustic Oscillations (BAO).}
\end{abstract}

@@

The hydrogen gas that dominates the baryon content of the Universe has undergone two phase transitions over cosmic history. The first was the transition from fully ionized to fully neutral gas during cosmic \emph{recombination\/} $\sim 300,000$ years after the Big Bang ($z\approx 1100$). This epoch has been quantified in exquisite detail through cosmic microwave background radiation studies \cite{planck15i}. The second transition is known as cosmic \emph{reionization}, when light from the first stars and black holes reionized the neutral Hydrogen (\HI) gas that pervaded the early Universe. Current indirect constraints suggest this second transition occurred a few hundred Myr to $\sim 1$~Gyr ($z\gtrsim 6$) \cite{gre17a} after the Big Bang. During this epoch, the photon-to-baryon ratio \replaced[id=NT]{rose significantly}{eventually exceeded unity}, reionizing the entire Universe except gas bound to galaxies. The astrophysical processes during this epoch shaped the evolution of galaxies and large-scale structure. However, unlike recombination, the timing and process of cosmic reionization remain poorly understood. The redshifted 21~cm spectral line from the electron spin-flip transition in \HI\ atom, has been recognized as the most promising tool to unravel the physical processes involved in cosmic reionization, and the evolution of large scale structure during the formation of the first galaxies \cite{sun72,sco90,mad97,toz00,ili02,fan02,fan06,bar07,mor10}.

Our Galaxy and intervening radio galaxies emit synchrotron radiation in the same frequency band ($\sim$~100--200~MHz) as the redshifted \HI\ signal from the epoch of reionization (EoR), with a brightness temperature roughly $\gtrsim 10^4$ times higher than the EoR \HI\ signal. However, the synchrotron foregrounds have a smooth spectrum whereas the EoR \HI\ signal will appear as small fluctuations superimposed on the smooth foreground spectrum. The hope of separating the cosmic line signal from the Galactic and extragalactic foregrounds has spawned a number of low-frequency instruments (eg. \cite{par10,tin13,van13}), with sensitivities sufficient for a statistical detection of the EoR \HI\ signal using power spectrum methodology \cite{thy13,bea13}.  However, despite obtaining \replaced[id=NT]{significant}{sufficient} observational data, these experiments remain dynamic range limited due to the coupling of the strong foreground emission to the EoR \HI\ signal via instrumental effects, particularly systematics related to precision of calibration, and the intrinsic chromatic instrumental response \cite{thy15a,thy15b,thy16,dat09,dat10,tro16, pac13,ali15,patil17,pob15,liu10,zhe14,barry16,sie17,dil17}.

@@

\added[id=NT]{Further, \HI\ intensity mapping experiments, such as CHIME \cite{ban14}, HIRAX \cite{new16}, and Tianlai \cite{xu15} will map \HI\ on large-scales at $1\lesssim z \lesssim 3$ tracing the matter density fluctuations, and thus measure the Baryon Acoustic Oscillations (BAO) that can be used as a standard ruler in constraining the {\it Dark Energy} equation of state. The challenges in these experiments are expected to be of a similar magnitude to that in 21~cm EoR experiments, and thus can potentially benefit from our approach.}

@@

The bi-spectrum in the context of interferometry has been investigated in \cite{jen58,kul89,tay99,tho01,mon06} and recently revisited in \cite{car18}. Consider a triad of antennas, $\{a, b, c\}$, indexed by $i$. $V_{ij}^\textrm{m}(f)$ is the spatial coherence measured at frequency, $f$, between antenna pairs, $\{ab, bc, ca\}$, indexed by $ij$. An interferometer measures $V_{ij}^\textrm{m}(f) = g_i(f)\, g_j^*(f)\, V_{ij}^\textrm{T}(f) + V_{ij}^\textrm{N}(f)$, the  true spatial coherence, $V_{ij}^\textrm{T}(f)$, corrupted by multiplicative antenna gains and ionospheric distortions, $g_i$, and additive thermal-like noise, $V_{ij}^\textrm{N}(f)$. \replaced[id=NT]{We}{It is pertinent to} express $V_{ij}^\textrm{T}(f) = V_{ij}^\textrm{F}(f) + V_{ij}^\textrm{H{\sc i}}(f)$, where, $V_{ij}^\textrm{F}(f)$ and $V_{ij}^\textrm{H{\sc i}}(f)$ are \replaced[id=NT]{due to}{contributions from} foreground synchrotron and EoR \HI\ signal, respectively. The measured bi-spectrum is $B_\nabla^\textrm{m}=\prod_{ij} V_{ij}^\textrm{m}=B_\nabla^\textrm{T}\prod_i |g_i|^2+B_\nabla^\textrm{N}$, where, the dependence on $f$ has been omitted for convenience (hereafter so, unless indicated). \added[id=NT]{In our notation, $ij$ only indexes three specific cyclic antenna pairs $\{ab, bc, ca\}$ and not all possible combinations of $i$ and $j$.} \deleted[id=NT]{The noise term,} $B_\nabla^\textrm{N}$, \added[id=NT]{encompassing all terms containing $V_{ij}^\textrm{N}$ including cross-terms,} has zero mean since it contains $V_{ij}^\textrm{N}$, \added[id=NT]{which is assumed to be uncorrelated between antenna pairs}. The phase of $B_\nabla^\textrm{m}$ is:
\begin{align}
  \phi_\nabla^\textrm{m} &= \sum_{ij}\phi_{ij}^\textrm{m} = \phi_\nabla^\textrm{T} + \delta\phi_\nabla^\textrm{N} = \left(\phi_\nabla^\textrm{F} + \delta\phi_\nabla^\textrm{H{\sc i}}\right) + \delta\phi_\nabla^\textrm{N} \label{eqn:cpphase-sum},
\end{align}
where, $\phi_{ij}^\textrm{m} = \phi_i - \phi_j + \phi_{ij}^\textrm{T} + \delta\phi_{ij}^\textrm{N}$, and $\phi_i$ are the phases of $V_{ij}^\textrm{m}$ and $g_i$, respectively. $\delta\phi_\nabla^\textrm{H{\sc i}}$ and $\delta\phi_\nabla^\textrm{N}$ \added[id=NT]{are not just determined by the phases of $V_{ij}^\textrm{H{\sc i}}$ and $V_{ij}^\textrm{N}$ but are small} \deleted[id=NT]{may be considered as small} perturbations due to \replaced[id=NT]{the EoR \HI\ signal and noise}{noise and EoR \HI\ signal} respectively, superimposed on dominant phase, $\phi_\nabla^\textrm{F}$, due to foregrounds, \added[id=NT]{and depend on $\rho_{ij}^\textrm{H{\sc i}}\equiv |V_{ij}^\textrm{F}|/|V_{ij}^\textrm{H{\sc i}}|$ and $\rho_{ij}^\textrm{N}\equiv |V_{ij}^\textrm{F}|/|V_{ij}^\textrm{N}|$ respectively.} \replaced[id=NT]{The small-perturbation formalism}{and} is valid when \replaced[id=NT]{$\rho_{ij}^\textrm{H{\sc i}}\gg 1$ and $\rho_{ij}^\textrm{N}\gg 1$}{$\rho_{ij}^\textrm{N} \equiv |V_{ij}^\textrm{F}|/|V_{ij}^\textrm{N}| \gg 1$ and $\rho_{ij}^\textrm{H{\sc i}} \equiv |V_{ij}^\textrm{F}|/|V_{ij}^\textrm{H{\sc i}}| \gg 1$}, as is usually the case in \replaced[id=NT]{low-frequency cosmology}{EoR \HI\ } experiments. $\phi_\nabla^\textrm{m}$ is independent of the antenna gains and measures $\phi_\nabla^\textrm{T}$, corrupted only by noise, $\delta\phi_\nabla^\textrm{N}$. Hence, $\phi_\nabla^\textrm{m}$ contains information purely about the sky's spatial structure, with certain geometric properties \cite[see][]{mon07}.

@@

\begin{figure}[htb]
\includegraphics[width=\linewidth]{visamp_spectra_asm_eor_noise}
\caption{Spatial coherence amplitude spectra on \replaced[id=NT]{antenna pairs forming a 14.6~m equilateral triad}{three non-redundant 14.6~m antenna spacings} are shown in red, blue and cyan. Solid \added[id=NT]{and dotted} lines show amplitudes from synchrotron foregrounds \added[id=NT]{and a fiducial EoR model, respectively}. Gray curve shows the typical noise obtained with 1~min integration. \deleted[id=NT]{Dotted curves show spatial coherence amplitudes from a fiducial EoR signal.} Typically, the foregrounds are $\gtrsim 10^4$ times brighter than the EoR signal. The orange bands denote the frequency sub-bands centered on 125~MHz, 150~MHz, and 175~MHz each of 10~MHz effective bandwidth. \label{fig:vis-spectra}}
\end{figure}

@@

Thus, when $\rho_{ij}^\textrm{H{\sc i}} < \rho_{ij}^\textrm{N}$, $\sigma_{\phi_\nabla^\textrm{m}}$ can be used to estimate $\rho_{ij}^\textrm{H{\sc i}}$, which in turn can be used to infer $|V_{ij}^\textrm{H{\sc i}}|$ if \replaced[id=NT]{$|V_{ij}^\textrm{F}|$ }{$V_{ij}^\textrm{F}$} is known. $|V_{ij}^\textrm{H{\sc i}}|$ is directly related to the EoR H\,{\sc i} brightness temperature fluctuations, $\langle(\delta T_\textrm{b}^\textrm{H{\sc i}})^2\rangle$, which can be inferred as a function of  redshift. This is discussed further below.

@@

For bi-spectrum phase, we consider 14.6~m equilateral antenna triads and assume ideal redundancy for the array. The total number of such measurements is assumed to be $N_\textrm{m} \sim 10^6$ (after allowing for $\sim$~50\% efficiency of data quality from $\sim 30$ redundant 14.6~m equilateral triads with HERA-47 \added[id=NT]{\cite[see layout in][]{car18}}, $\sim$~8~hours of observing per day at 1~min integration intervals, repeated over $\sim$~150 nights). We assume there are $N_\textrm{c}$ coherent measurements of $\phi_\nabla^\textrm{m}$ and for each of these measurements, there are $N_\textrm{ic}$ incoherent measurements such that $N_\textrm{c}N_\textrm{ic}=N_\textrm{m}$. $\phi_\nabla^\textrm{m}$ is averaged coherently over $N_\textrm{c}$ measurements, and then averaged incoherently in $P_\nabla(k_\parallel)$ measured for all incoherent pairs of $\phi_\nabla^\textrm{m}$. We choose $W(f)$ to be the inverse Fourier transform of the squared delay response of a {\it Blackman-Harris} spectral weighting \cite{har78} as proposed in \cite{thy16}, on sub-bands centered at 125~MHz, 150~MHz, and 175~MHz, with an effective bandwidth of $B_\textrm{eff}\simeq 10$~MHz (to minimize EoR signal evolution over the redshift range).

@@

\begin{figure*}[htb]
  \includegraphics[width=\linewidth]{cpdps_all_subbands_SNR_200_nsamples_1048576}
\caption{Delay power spectra of bi-spectrum phases in sub-bands centered at 125~MHz (left), 150~MHz (middle), and 175~MHz (right) that include contributions from foregrounds, EoR signal, and noise. \added[id=NT]{The corresponding redshifts are listed in each panel. The 14.6~m equilateral triad used is highlighted on a HERA-19 layout on the top left.} \replaced[id=NT]{Green curve shows only the foreground contribution ($\phi_\nabla^\textrm{m}=\phi_\nabla^\textrm{F}$).}{The orange curve applies when only foregrounds are present, $\phi_\nabla^\textrm{m}=\phi_\nabla^\textrm{F}$.} The anisotropic foreground model employed \cite{thy15a} results in the asymmetry around $k_\parallel=0\,h^{-1}$~Mpc. The red curve \replaced[id=NT]{shows the effect of including}{is when} noise fluctuations \replaced[id=NT]{($\phi_\nabla^\textrm{m}=\phi_\nabla^\textrm{F}+\delta\phi_\nabla^\textrm{N}$)}{for noise fluctuations, $\delta\phi_\nabla^\textrm{N}$ superimposed on the foreground component, $\phi_\nabla^\textrm{F}$}, after averaging $N_\textrm{m} \sim 10^6$ measurements. The gray curve is when fluctuations from the EoR \HI\ signal are \replaced[id=NT]{included}{also present}, $\phi_\nabla^\textrm{m}=\phi_\nabla^\textrm{F}+\delta\phi_\nabla^\textrm{H{\sc i}}+\delta\phi_\nabla^\textrm{N}$. The black curve is \replaced[id=NT]{identical to}{the same as} the gray one except the EoR \HI\ signal is 10 times brighter. The dot and vertical markers denote positive and negative values respectively. The latter results from the removal of noise bias by using cross-power spectra (Eq.~\ref{eqn:delay-power-spectrum}). The EoR \HI\ contribution is significantly detected at $|k_\parallel| \gtrsim 0.5\,h$~Mpc$^{-1}$ at $z\approx 8.5$ and $z\approx 7.1$. At $z\approx 10.4$, the EoR contribution is \deleted[id=NT]{virtually} inseparable from \deleted[id=NT]{the} foregrounds because \replaced[id=NT]{the EoR signal is weaker (see power spectrum of {\sc faint galaxies} model at $k=0.5$~Mpc$^{-1}$ in Fig.~2 of \cite{gre17b}) while the foregrounds are brighter}{they are brighter and the EoR signal is weaker (see power spectrum of {\sc faint galaxies} model at $k=0.5$~Mpc$^{-1}$ in Fig.~2 of \cite{gre17b})}. The detection significance depends on the strength of EoR H\,{\sc i} signal relative to the foregrounds. A 10-fold increase in EoR intensity shows a $\sim 100$-fold increase in $P_\nabla(k_\parallel)$ at $|k_\parallel| \gtrsim 0.5\,h$~Mpc$^{-1}$ at $z\approx 8.5$ and $z\approx 7.1$, indicating its sensitive dependence on $\rho_{ij}^\textrm{H{\sc i}}$. This detection measures the line-to-continuum ratio, from which the EoR H\,{\sc i} brightness temperature fluctuations, $\langle(\delta T_\textrm{b}^\textrm{H{\sc i}}(z))^2\rangle$, can be inferred using a reliable foreground model. \label{fig:cpdps}}
\end{figure*}

>>>>

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{amssymb}   % for math
\usepackage{hyperref}% add hypertext capabilities
\usepackage{aas_macros}
\usepackage[caption=false]{subfig}
\usepackage[draft]{changes}
% \definechangesauthor[name={Nithyanandan Thyagarajan}, color=red]{NT}
% \setremarkmarkup{(#2)}
\colorlet{Changes@Color}{red}

% \usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\begin{abstract} 
  Detecting neutral Hydrogen (H\,{\sc i}) via the 21~cm line emission from the intergalactic medium at $z\gtrsim 6$ has been identified as one of the most promising probes of the epoch of cosmic reionization -- a major phase transition of the Universe. However, these studies face severe challenges imposed by the bright foreground emission from cosmic objects. Current techniques require precise instrumental calibration to separate the weak H\,{\sc i} line signal from the foreground continuum emission. We propose to mitigate this calibration requirement by using \replaced{measurements of the interferometric bi-spectrum phase}{the phase of the bi-spectrum from interferometric measurements}. Bi-spectrum phase is unaffected by antenna-based direction-independent calibration errors and hence for a compact array it depends on the sky brightness distribution only (subject to the usual thermal-like noise). We show that the bi-spectrum phase of foreground synchrotron continuum has a characteristically smooth spectrum relative to the cosmological line signal. The two can be separated effectively by exploiting this spectral difference using Fourier techniques, while eliminating the need for precise antenna-based calibration of phases introduced by the instrument, and the ionosphere, inherent in existing approaches. Using fiducial models for continuum foregrounds, and for the cosmological H\,{\sc i} signal, we show the latter should be detectable in bi-spectrum phase spectra, with reasonable significance at $|k_\parallel| \gtrsim 0.5\,h$~Mpc$^{-1}$, using existing instruments. \added{Our approach will also benefit other H\,{\sc i} intensity mapping experiments that face similar challenges, such as those measuring Baryon Acoustic Oscillations (BAO).}
\end{abstract}

@@

The hydrogen gas that dominates the baryon content of the Universe has undergone two phase transitions over cosmic history. The first was the transition from fully ionized to fully neutral gas during cosmic \emph{recombination\/} $\sim 300,000$ years after the Big Bang ($z\approx 1100$). This epoch has been quantified in exquisite detail through cosmic microwave background radiation studies \cite{planck15i}. The second transition is known as cosmic \emph{reionization}, when light from the first stars and black holes reionized the neutral Hydrogen (\HI) gas that pervaded the early Universe. Current indirect constraints suggest this second transition occurred a few hundred Myr to $\sim 1$~Gyr ($z\gtrsim 6$) \cite{gre17a} after the Big Bang. During this epoch, the photon-to-baryon ratio \replaced{rose significantly}{eventually exceeded unity}, reionizing the entire Universe except gas bound to galaxies. The astrophysical processes during this epoch shaped the evolution of galaxies and large-scale structure. However, unlike recombination, the timing and process of cosmic reionization remain poorly understood. The redshifted 21~cm spectral line from the electron spin-flip transition in \HI\ atom, has been recognized as the most promising tool to unravel the physical processes involved in cosmic reionization, and the evolution of large scale structure during the formation of the first galaxies \cite{sun72,sco90,mad97,toz00,ili02,fan02,fan06,bar07,mor10}.

Our Galaxy and intervening radio galaxies emit synchrotron radiation in the same frequency band ($\sim$~100--200~MHz) as the redshifted \HI\ signal from the epoch of reionization (EoR), with a brightness temperature roughly $\gtrsim 10^4$ times higher than the EoR \HI\ signal. However, the synchrotron foregrounds have a smooth spectrum whereas the EoR \HI\ signal will appear as small fluctuations superimposed on the smooth foreground spectrum. The hope of separating the cosmic line signal from the Galactic and extragalactic foregrounds has spawned a number of low-frequency instruments (eg. \cite{par10,tin13,van13}), with sensitivities sufficient for a statistical detection of the EoR \HI\ signal using power spectrum methodology \cite{thy13,bea13}.  However, despite obtaining \replaced{significant}{sufficient} observational data, these experiments remain dynamic range limited due to the coupling of the strong foreground emission to the EoR \HI\ signal via instrumental effects, particularly systematics related to precision of calibration, and the intrinsic chromatic instrumental response \cite{thy15a,thy15b,thy16,dat09,dat10,tro16, pac13,ali15,patil17,pob15,liu10,zhe14,barry16,sie17,dil17}.

@@

\added{Further, \HI\ intensity mapping experiments, such as CHIME \cite{ban14}, HIRAX \cite{new16}, and Tianlai \cite{xu15} will map \HI\ on large-scales at $1\lesssim z \lesssim 3$ tracing the matter density fluctuations, and thus measure the Baryon Acoustic Oscillations (BAO) that can be used as a standard ruler in constraining the {\it Dark Energy} equation of state. The challenges in these experiments are expected to be of a similar magnitude to that in 21~cm EoR experiments, and thus can potentially benefit from our approach.}

@@

The bi-spectrum in the context of interferometry has been investigated in \cite{jen58,kul89,tay99,tho01,mon06} and recently revisited in \cite{car18}. Consider a triad of antennas, $\{a, b, c\}$, indexed by $i$. $V_{ij}^\textrm{m}(f)$ is the spatial coherence measured at frequency, $f$, between antenna pairs, $\{ab, bc, ca\}$, indexed by $ij$. An interferometer measures $V_{ij}^\textrm{m}(f) = g_i(f)\, g_j^*(f)\, V_{ij}^\textrm{T}(f) + V_{ij}^\textrm{N}(f)$, the  true spatial coherence, $V_{ij}^\textrm{T}(f)$, corrupted by multiplicative antenna gains and ionospheric distortions, $g_i$, and additive thermal-like noise, $V_{ij}^\textrm{N}(f)$. \replaced{We}{It is pertinent to} express $V_{ij}^\textrm{T}(f) = V_{ij}^\textrm{F}(f) + V_{ij}^\textrm{H{\sc i}}(f)$, where, $V_{ij}^\textrm{F}(f)$ and $V_{ij}^\textrm{H{\sc i}}(f)$ are \replaced{due to}{contributions from} foreground synchrotron and EoR \HI\ signal, respectively. The measured bi-spectrum is $B_\nabla^\textrm{m}=\prod_{ij} V_{ij}^\textrm{m}=B_\nabla^\textrm{T}\prod_i |g_i|^2+B_\nabla^\textrm{N}$, where, the dependence on $f$ has been omitted for convenience (hereafter so, unless indicated). \added{In our notation, $ij$ only indexes three specific cyclic antenna pairs $\{ab, bc, ca\}$ and not all possible combinations of $i$ and $j$.} \deleted{The noise term,} $B_\nabla^\textrm{N}$, \added{encompassing all terms containing $V_{ij}^\textrm{N}$ including cross-terms,} has zero mean since it contains $V_{ij}^\textrm{N}$, \added{which is assumed to be uncorrelated between antenna pairs}. The phase of $B_\nabla^\textrm{m}$ is:
\begin{align}
  \phi_\nabla^\textrm{m} &= \sum_{ij}\phi_{ij}^\textrm{m} = \phi_\nabla^\textrm{T} + \delta\phi_\nabla^\textrm{N} = \left(\phi_\nabla^\textrm{F} + \delta\phi_\nabla^\textrm{H{\sc i}}\right) + \delta\phi_\nabla^\textrm{N} \label{eqn:cpphase-sum},
\end{align}
where, $\phi_{ij}^\textrm{m} = \phi_i - \phi_j + \phi_{ij}^\textrm{T} + \delta\phi_{ij}^\textrm{N}$, and $\phi_i$ are the phases of $V_{ij}^\textrm{m}$ and $g_i$, respectively. $\delta\phi_\nabla^\textrm{H{\sc i}}$ and $\delta\phi_\nabla^\textrm{N}$ \added{are not just determined by the phases of $V_{ij}^\textrm{H{\sc i}}$ and $V_{ij}^\textrm{N}$ but are small} \deleted{may be considered as small} perturbations due to \replaced{the EoR \HI\ signal and noise}{noise and EoR \HI\ signal} respectively, superimposed on dominant phase, $\phi_\nabla^\textrm{F}$, due to foregrounds, \added{and depend on $\rho_{ij}^\textrm{H{\sc i}}\equiv |V_{ij}^\textrm{F}|/|V_{ij}^\textrm{H{\sc i}}|$ and $\rho_{ij}^\textrm{N}\equiv |V_{ij}^\textrm{F}|/|V_{ij}^\textrm{N}|$ respectively.} \replaced{The small-perturbation formalism}{and} is valid when \replaced{$\rho_{ij}^\textrm{H{\sc i}}\gg 1$ and $\rho_{ij}^\textrm{N}\gg 1$}{$\rho_{ij}^\textrm{N} \equiv |V_{ij}^\textrm{F}|/|V_{ij}^\textrm{N}| \gg 1$ and $\rho_{ij}^\textrm{H{\sc i}} \equiv |V_{ij}^\textrm{F}|/|V_{ij}^\textrm{H{\sc i}}| \gg 1$}, as is usually the case in \replaced{low-frequency cosmology}{EoR \HI\ } experiments. $\phi_\nabla^\textrm{m}$ is independent of the antenna gains and measures $\phi_\nabla^\textrm{T}$, corrupted only by noise, $\delta\phi_\nabla^\textrm{N}$. Hence, $\phi_\nabla^\textrm{m}$ contains information purely about the sky's spatial structure, with certain geometric properties \cite[see][]{mon07}.

@@

\begin{figure}[htb]
\includegraphics[width=\linewidth]{visamp_spectra_asm_eor_noise}
\caption{Spatial coherence amplitude spectra on \replaced{antenna pairs forming a 14.6~m equilateral triad}{three non-redundant 14.6~m antenna spacings} are shown in red, blue and cyan. Solid \added{and dotted} lines show amplitudes from synchrotron foregrounds \added{and a fiducial EoR model, respectively}. Gray curve shows the typical noise obtained with 1~min integration. \deleted{Dotted curves show spatial coherence amplitudes from a fiducial EoR signal.} Typically, the foregrounds are $\gtrsim 10^4$ times brighter than the EoR signal. The orange bands denote the frequency sub-bands centered on 125~MHz, 150~MHz, and 175~MHz each of 10~MHz effective bandwidth. \label{fig:vis-spectra}}
\end{figure}

@@

Thus, when $\rho_{ij}^\textrm{H{\sc i}} < \rho_{ij}^\textrm{N}$, $\sigma_{\phi_\nabla^\textrm{m}}$ can be used to estimate $\rho_{ij}^\textrm{H{\sc i}}$, which in turn can be used to infer $|V_{ij}^\textrm{H{\sc i}}|$ if \replaced{$|V_{ij}^\textrm{F}|$ }{$V_{ij}^\textrm{F}$} is known. $|V_{ij}^\textrm{H{\sc i}}|$ is directly related to the EoR H\,{\sc i} brightness temperature fluctuations, $\langle(\delta T_\textrm{b}^\textrm{H{\sc i}})^2\rangle$, which can be inferred as a function of  redshift. This is discussed further below.

@@

For bi-spectrum phase, we consider 14.6~m equilateral antenna triads and assume ideal redundancy for the array. The total number of such measurements is assumed to be $N_\textrm{m} \sim 10^6$ (after allowing for $\sim$~50\% efficiency of data quality from $\sim 30$ redundant 14.6~m equilateral triads with HERA-47 \added{\cite[see layout in][]{car18}}, $\sim$~8~hours of observing per day at 1~min integration intervals, repeated over $\sim$~150 nights). We assume there are $N_\textrm{c}$ coherent measurements of $\phi_\nabla^\textrm{m}$ and for each of these measurements, there are $N_\textrm{ic}$ incoherent measurements such that $N_\textrm{c}N_\textrm{ic}=N_\textrm{m}$. $\phi_\nabla^\textrm{m}$ is averaged coherently over $N_\textrm{c}$ measurements, and then averaged incoherently in $P_\nabla(k_\parallel)$ measured for all incoherent pairs of $\phi_\nabla^\textrm{m}$. We choose $W(f)$ to be the inverse Fourier transform of the squared delay response of a {\it Blackman-Harris} spectral weighting \cite{har78} as proposed in \cite{thy16}, on sub-bands centered at 125~MHz, 150~MHz, and 175~MHz, with an effective bandwidth of $B_\textrm{eff}\simeq 10$~MHz (to minimize EoR signal evolution over the redshift range).

@@

\begin{figure*}[htb]
  \includegraphics[width=\linewidth]{cpdps_all_subbands_SNR_200_nsamples_1048576}
\caption{Delay power spectra of bi-spectrum phases in sub-bands centered at 125~MHz (left), 150~MHz (middle), and 175~MHz (right) that include contributions from foregrounds, EoR signal, and noise. \added{The corresponding redshifts are listed in each panel. The 14.6~m equilateral triad used is highlighted on a HERA-19 layout on the top left.} \replaced{Green curve shows only the foreground contribution ($\phi_\nabla^\textrm{m}=\phi_\nabla^\textrm{F}$).}{The orange curve applies when only foregrounds are present, $\phi_\nabla^\textrm{m}=\phi_\nabla^\textrm{F}$.} The anisotropic foreground model employed \cite{thy15a} results in the asymmetry around $k_\parallel=0\,h^{-1}$~Mpc. The red curve \replaced{shows the effect of including}{is when} noise fluctuations \replaced{($\phi_\nabla^\textrm{m}=\phi_\nabla^\textrm{F}+\delta\phi_\nabla^\textrm{N}$)}{for noise fluctuations, $\delta\phi_\nabla^\textrm{N}$ superimposed on the foreground component, $\phi_\nabla^\textrm{F}$}, after averaging $N_\textrm{m} \sim 10^6$ measurements. The gray curve is when fluctuations from the EoR \HI\ signal are \replaced{included}{also present}, $\phi_\nabla^\textrm{m}=\phi_\nabla^\textrm{F}+\delta\phi_\nabla^\textrm{H{\sc i}}+\delta\phi_\nabla^\textrm{N}$. The black curve is \replaced{identical to}{the same as} the gray one except the EoR \HI\ signal is 10 times brighter. The dot and vertical markers denote positive and negative values respectively. The latter results from the removal of noise bias by using cross-power spectra (Eq.~\ref{eqn:delay-power-spectrum}). The EoR \HI\ contribution is significantly detected at $|k_\parallel| \gtrsim 0.5\,h$~Mpc$^{-1}$ at $z\approx 8.5$ and $z\approx 7.1$. At $z\approx 10.4$, the EoR contribution is \deleted{virtually} inseparable from \deleted{the} foregrounds because \replaced{the EoR signal is weaker (see power spectrum of {\sc faint galaxies} model at $k=0.5$~Mpc$^{-1}$ in Fig.~2 of \cite{gre17b}) while the foregrounds are brighter}{they are brighter and the EoR signal is weaker (see power spectrum of {\sc faint galaxies} model at $k=0.5$~Mpc$^{-1}$ in Fig.~2 of \cite{gre17b})}. The detection significance depends on the strength of EoR H\,{\sc i} signal relative to the foregrounds. A 10-fold increase in EoR intensity shows a $\sim 100$-fold increase in $P_\nabla(k_\parallel)$ at $|k_\parallel| \gtrsim 0.5\,h$~Mpc$^{-1}$ at $z\approx 8.5$ and $z\approx 7.1$, indicating its sensitive dependence on $\rho_{ij}^\textrm{H{\sc i}}$. This detection measures the line-to-continuum ratio, from which the EoR H\,{\sc i} brightness temperature fluctuations, $\langle(\delta T_\textrm{b}^\textrm{H{\sc i}}(z))^2\rangle$, can be inferred using a reliable foreground model. \label{fig:cpdps}}
\end{figure*}
